---
title: "Introduction"
author: "Rich FitzJohn"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

``` {r echo=FALSE, results="hide"}
local({
  path <- tempfile()
  cyphr::ssh_keygen(path, FALSE)
  Sys.setenv(USER_KEY=path)
})
```

This package tries to smooth over some of the differences in
encryption approaches (symmetric vs asymmetric, sodium vs openssl)
to provide a simple interface for users who just want to encrypt or
decrypt things.  My main motivation for writing this package is
described in the [data workflow](data.html) vignette but
implementing that requires a few features that might be generally
useful.

This vignette works through the basic functionality of the package.
It does not offer much in the way of an introduction to encryption
itself; for that see the excellent vignettes in the `openssl` and
`sodium` packages.  This package is a thin wrapper for those
packages.

# Keys and the like

To encrypt anything we need a key.  There are two sorts of key
"types" we will concern ourselves with here "symmetric" and
"asymmetric".

* "symmetric" keys are used for storing secrets that multiple
  people need to access.  Everyone has the same key (which is just
  a bunch of bytes) and with that we can either encrypt data or
  decrypt it.

* a "key pair" is a public and a private key; this is used in
  communication.  You hold a private key that nobody else ever sees
  and a public key that you can copy around all over the show.
  These can be used for a couple of different patterns of
  communication (see below).

We support key pairs from `openssl` and from `sodium` and symmetric
keys from `sodium`.  In the rest of the package I use key pair from
`openssl` to encrypt a `sodium` symmetric key.  The reason for this
is that openssl keys have a common file format and many people have
one already (whereas I do not see that `sodium` has anything like
this).  On the other hand, `sodium` is really fast modern so it's
nice to use that for dealing with potentially large amounts of
data.

## Symmetric keys (`sodium`)

A `sodium` key is literally just 32 bytes of random data.
``` {r }
key <- sodium::keygen()
key
```

Because the key is not readily idenfiable as a key we wrap it up
with \code{config_sodium_symmetric} (all of the approaches will
work this way).
``` {r }
cfg <- cyphr::config_sodium_symmetric(key)
```

With this we can use the encryption functions in the package (see below)

## Asymmetric keys (`openssl`)

With `openssl` you have a key pair and can do "public key
encryption".  With that, anyone with a copy of your public key can
send you a message that only you, with your private key, can
decrypt.

To load your key pair, including your private key, you could run:
``` {r }
pair <- cyphr::config_openssl()
```

which will look, in turn, at:

* the argument `path` (`NULL` by default) to `config_openssl`
* the environment variable (`USER_KEY` or `USER_PUBKEY`)
* the path `~/.ssh/id_rsa.pub`

The path provided can be the directory containing `id_rsa.pub` and
`id_rsa` or a path to either and hopefully the right thing will
happen.  If your private key is password-protected (a *very* good
idea) you will be prompted for your password.

Someone wanting to send you a message would load your public key with:

```r
pair <- cyphr::config_openssl(path, private=FALSE)
```

which would load your public key from `path` and _not_ try to load
a private key.

## Asymmetric keys (`sodium`)

This is currently less useful practically because there is no
standard format for storing the keys.  However, you can save them
to a file with `writeBin` and `cyphr` will read them from there.
Alternatively if you have the raw bytes of the keys (which is what
`sodium` returns) pass those in.
``` {r }
key <- sodium::keygen()
pub <- sodium::pubkey(key)
key
pub
```

With sodium key pairs there are two sorts of encryption one can do
(see the sodium vignette for more details here).  First, and most
similarly to the `openss` method, we can do public key encryption
by using `cyphr::config_sodium_public`:
``` {r }
pair <- cyphr::config_sodium_public(pub, key)
```

Note that ordinarily you would use someone elses public key here!

Second, you could do _authenticated_ public key encryption with
``` {r }
pair <- cyphr::config_sodium_authenticated(pub, key)
```

With this approach both the sender and recipient need the to know
each others public keys.  With public key encryption _anyone_ can
send a message to a recipient but with this approach only this pair
of people can communicate.  See `?sodium::auth_encrypt` for more
details.

# Encrypting things

## Files, via the high level interface

This is the most user-friendly way of using the package.  The
package provides a pair of functions `encrypt` and `decrypt` that
wrap file writing and file reading functions.  In general you would
use `encrypt` when writing a file and `decrypt` when reading one.
They're designed to be used like so:

Suppose you have a super-secret object that you want to share privately
``` {r }
cfg <- cyphr::config_sodium_symmetric(sodium::keygen())
x <- list(a = 1:10, b = "don't tell anyone else")
```

If you save this to disk with `saveRDS` it will be readable by
everyone.  But if you encrypted the file that `saveRDS` produced it
would be protected:
``` {r }
cyphr::encrypt(saveRDS(x, "secret.rds"), cfg)
```

(see below for some more details on how this works).

This file cannot be read with `readRDS`:

``` {r error=TRUE}
readRDS("secret.rds")
```

but if we wrap the call with `decrypt` and pass in the config
object it can be decrypted and read:
``` {r }
cyphr::decrypt(readRDS("secret.rds"), cfg)
```

What happens in the call above is some moderately nasty call
rewriting.  If this bothers you, you should just use `encrypt_file`
/ `decrypt_file` and make sure to clean up after yourself.

The `encrypt` function inspects the call in the first argument
passed to it and works out for the function provided (`saveRDS`)
which argument corresponds to the filename (here `"secret.rds"`).
It then rewrites the call to write out to a temporary file (using
`tempfile()`).  Then it calls `encrypt_file` (see below) on this
temporary file to create the file asked for (`"secret.rds"`).  Then
it deletes the temporary file, though this will also happen in case
of an error in any of the above.

The `decrypt` function works similarly.  It insepects the call and
detects that the first argument represents the filename.  It
decrypts that file to create a temporary file, and then runs
`readRDS` on that file.  Again it will delete the temporary file on
exit.

The functions supported via this interface are:

* `readLines` / `writeLines`
* `readRDS` / `writeRDS`
* `read` / `save`
* `read.table` / `write.table`
* `read.csv` / `read.csv2` / `write.csv`
* `read.delim` / `read.delim2`

But new functions can be added with the `rewrite_register`
function.  For example, to support the excellent
[rio](https://cran.r-project.org/web/packages/rio) package, whose
`import` and `export` functions take the filename `file` you could
use:

```r
cyphr::rewrite_register("rio", "import", "file")
cyphr::rewrite_register("rio", "export", "file")
```

now you can read and write tabular data into and out of a great
many different file formats with encryption with calls like

```r
cyphr::encrypt(rio::export(mtcars, "file.json"), cfg)
cyphr::decrypt(rio::import("file.json"), cfg)
```

The functions above use [non standard evaluation](http://adv-r...)
and so may not be suitable for programming or use in packages.  An
"escape hatch" is provided via `encrypt_` and `decrypt_` where the
first argument is a quoted expression.
``` {r }
cyphr::encrypt_(quote(saveRDS(x, "secret.rds")), cfg)
cyphr::decrypt_(quote(readRDS("secret.rds")), cfg)
```

## Objects
``` {r }
cfg <- cyphr::config_sodium_symmetric(sodium::keygen())
```

Here's an object to encrypt:
``` {r }
obj <- list(x=1:10, y="secret")
```

This creates a bunch of raw bytes corresponding to the data (it's
not really possible to print this as anything nicer than bytes).
``` {r }
secret <- cyphr::encrypt_object(obj, NULL, cfg)
secret
```

(the `NULL` second argument to the `encrypt_object` call indicates
we don't want to write to file but instead want to return the data
itself, like R's `serialize` function).

The data can be decrypted with the `decrypt_object` function:
``` {r }
cyphr::decrypt_object(secret, cfg)
```

## Strings

For the case of strings we can do this in a slightly more
lightweight way (the above function routes through `serialize` /
`deserialize` which can be slow and will create larger objects than
using `charToRaw` / `rawToChar`)
``` {r }
secret <- cyphr::encrypt_string("secret", NULL, cfg)
secret
```

and decrypt:
``` {r }
cyphr::decrypt_string(secret, cfg)
```

## Plain raw data

If these are not enough for you, you can work directly with raw
objects (bunches of bytes) by using `encrypt_data`:
``` {r }
dat <- sodium::random(100)
dat # some random bytes

secret <- cyphr::encrypt_data(dat, NULL, cfg)
secret
```

Decrypted data is the same as a the original data
``` {r }
identical(cyphr::decrypt_data(secret, NULL, cfg), dat)
```

## Files

## Send a secret message to someone

Because it's not totally obvious, here is how two users can use
openssl rsa keypairs to share a secret key.  This is the idea
underlying the [workflow discussed elsewhere](data.html).

First, generate a keypair.
``` {r }
path_pair <- cyphr::ssh_keygen(tempfile(), FALSE)
pair <- cyphr::config_openssl(path_pair)
```

Next we can load the *public key* from the pair above.  Assume that
the public key has been swapped around by the users by this point,
and that the code here is being run on a different users's machine.
``` {r }
pub <- cyphr::config_openssl(path_pair, private=FALSE)
```

Then we'll generate a sodium symmetric key to share.
Alternatively, this could be any data.
``` {r }
data <- sodium::keygen()
```

Then we can use the public key of user2 to encrypt a message that
only they can read:
``` {r }
secret <- cyphr::encrypt_data(data, NULL, pub)
```

The user can then decrypt the data this way:
``` {r }
cyphr::decrypt_data(secret, NULL, pair)
```

Nobody else but that user can decrypt the data.

## Secure message exchange

If two users both know each others public keys (and their own
private keys) they can securely message each other.  This means
that the first user will encrypt a message with their private key
and the other users' public key and only the other user can decrypt
the message _and_ they know it has come from the first user.  This
is not currently supported but will be available by passing the
public and private keys explicitly to `config_openssl` in a future
version.

First generate two sets of keys:
``` {r }
key1 <- sodium::keygen()
pub1 <- sodium::pubkey(key1)

key2 <- sodium::keygen()
pub2 <- sodium::pubkey(key2)
```

The first user will create an authenticated config object using
_their_ private key and the _other users_ public key (and the
second user will do the opposite).
``` {r }
cfg1 <- cyphr::config_sodium_authenticated(pub2, key1)
cfg2 <- cyphr::config_sodium_authenticated(pub1, key2)
```

The first user can now encrypt a message:
``` {r }
x <- runif(10)
secret <- cyphr::encrypt_object(x, NULL, cfg1)
```

and the second user can decrypt it:
``` {r }
identical(cyphr::decrypt_object(secret, cfg2), x)
```

Now, suppose there is a third, malicious, user:
``` {r }
key3 <- sodium::keygen()
pub3 <- sodium::pubkey(key3)
```

as with public key encryption, they cannot decrypt the message
because it was not encrypted with their public key:

``` {r error=TRUE}
cyphr::decrypt_object(secret,
                         cyphr::config_sodium_authenticated(pub1, key3))
```

But unlike public key encryption they cannot send a fake message to
the second user.  Because the second user is using the first user's
public key, they expect a message _only_ from the first user:
``` {r }
cfg3 <- cyphr::config_sodium_authenticated(pub3, key1)
x3 <- x + 1 # malicious message
secret3 <- cyphr::encrypt_object(x3, NULL, cfg3)
```

When the second user tries to decrypt the message they get an error
``` {r error=TRUE}
cyphr::decrypt_object(secret3, cfg2)
```

``` {r echo=FALSE, results="hide"}
Sys.unsetenv("USER_KEY")
file.remove("secret.rds")
```
