---
title: "Data Encryption"
author: "Rich FitzJohn"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Encryption}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

``` {r echo=FALSE, results="hide"}
options(encryptr.user.path=tempfile())
unlink("data", recursive=TRUE)

library(encryptr)
```

The scenario:

A group of people are working on a sensitive data set that for
practical reasons needs to be stored in a place that we're not 100%
happy with the securtity (e.g., dropbox).  If the data can be
stored encrypted but everyone in the group can still read and write
the data then we've improved the situation somewhat.

The general proceedure is this:

1. A person will set up a set of personal keys and a key for the
data.  The data key will be encrypted with their personal key so
they have access to the data but nobody else does.  At this point
the data can be encrypted.

2. Additional users set up personal keys and request access to the
data.  Anyone with access to the data can grant access to anyone
else.

We'll store data in the directory `data`; at present there is
nothing there.
``` {r }
data_dir <- "data"
dir.create(data_dir, FALSE)
```

**First**, create a set of keys.  These will be shared across all
projects and stored away from the data.  This command is safe to
run repeatedly and will not overwrite any keys.
``` {r }
data_user_init()
```

You won't need to directly read the key often, but you can do so
with the `data_key_read` function, which at least shows what is
stored there.  There's a bit of personal information to make it a
little easier to see who a key belongs to.
``` {r }
data_key_read()
```

Your key has a "hash" associated with it that will appear in a few
places; to get it run:
``` {r }
data_key_read()$hash
```

You will need to run this command on each computer you want access
from.  Don't copy the key around.  Note that `data_user_init` will
be run automatically by subsequent commands (see the minimal
example below).

**Second**, create a key for the data and encrypt that key with
your personal key.
``` {r }
data_admin_init(data_dir)
```

Again, this command can be run multiple times safely.
``` {r }
data_admin_init(data_dir)
```

**Third**, you can add encrypted data to the directory (or to
anywhere really).  When run, `config_data` will verify that it can
actually decryt things.
``` {r }
cfg <- config_data(data_dir)
```

This object can be used with all the `encryptr` functions:
``` {r }
filename <- file.path(data_dir, "iris.rds")
encrypt(saveRDS(iris, filename), cfg)
dir(data_dir)
```

The file is encrypted:
``` {r error=TRUE}
readRDS(filename)
```

But we can decrypt it:
``` {r }
head(decrypt(readRDS(filename), cfg))
```

**Fourth**, have someone else join in.  To simulate another person
here, I'm going to pass an argument `user2` though.  If run on an
actually dfifferent computer this would not be needed; this is just
to simulate two users in a single session for this vignette (see
minimal example below).
``` {r }
user2 <- tempfile()
data_user_init(path=user2)
```

We're going to assume that the user can read and write to the data.
This is the case for my use case where the data are stored on
dropbox and will be the case with github based distribution, though
there would be a pull request step in here.

This user cannot read the data:
``` {r error=TRUE}
cfg2 <- config_data(data_dir, user2)
```

Write a hopeful key here and see we get denied access:
``` {r error=TRUE}
tmp <- file.path(data_dir, ".encryptr",
                 paste0(paste0(data_key_read(user2)$hash, collapse=""), ".key"))
writeBin(c(0, sodium::random(32)), tmp)
cfg2 <- config_data(data_dir, user2)
file.remove(tmp)
```

So that's positive; we have a data set that seems safe to the
lamest attack in the world.

But `user2` is your collaborator and needs access.  What they need
to do is run:
``` {r }
data_request_access(data_dir, user2)
```

The user should the send an email to someone with access and quote
the hash in the message above.

**Fifth**, back on the first computer we can authorise the second
user.  First, see who has requested access:
``` {r }
req <- data_admin_list_requests(data_dir)
req
```

We can see the same hash here as above (``r names(req)[[1]]``)

...and then grant access to them with the `data_admin_authorise` function.
``` {r }
data_admin_authorise(names(req)[[1]], data_dir)
```

which has cleared the request queue:
``` {r }
data_admin_list_requests(data_dir)
```

and added it to our set of keys:
``` {r }
data_admin_list_keys(data_dir)
```

**Finally**, as soon as the authorisation has happened, the user
can encrypt and decrypt files:
``` {r }
cfg2 <- config_data(data_dir, user2)
head(decrypt(readRDS(filename), cfg2))
```

## Minimal example

As above, but with less discussion:

``` {r echo=FALSE, results="hide"}
unlink(data_dir, recursive=TRUE)
unlink(user2, recursive=TRUE)
unlink(getOption("encryptr.user.path"), recursive=TRUE)
dir.create(data_dir)
```

Setup, on computer 1:
``` {r }
data_admin_init(data_dir)
```

Encrypt a file:
``` {r }
encrypt(saveRDS(iris, filename), config_data(data_dir))
```

Request access, on computer 2:
``` {r echo=FALSE}
oo <- options(encryptr.user.path=user2)
```
``` {r echo=TRUE}
hash <- data_request_access(data_dir)
```
``` {r echo=FALSE}
options(oo)
```

Authorise, on computer 1:
``` {r }
data_admin_authorise(hash, data_dir)
```

Read data, on computer 2:
``` {r echo=FALSE}
oo <- options(encryptr.user.path=user2)
```
``` {r echo=TRUE}
head(decrypt(readRDS(filename), config_data(data_dir)))
```
``` {r echo=FALSE}
options(oo)
```

## Details

Each user has a directory with a public and private key in it.  The
private key is just 32 bytes of binary data, and the public key is
a human-readable file with a little additional metadata to make it
easier for people to associate keys with people.

The data directory will have a hidden directory `.encryptr` in
it. This does not actually have to be stored with the data but it
makes sense to.  This directory will contain a set of encrypted
keys; these keys belong to different users and can be decrpted
using their private keys.

When an authorised user creates the `config_data` object, the package:

* reads their encrypted key from the data directory
* reads their private key from their user directory
* decrypts the data key to give a "symmetric" key which will be the
  same for all users (but is never directly stored anywhere).

Public keys are stored in the requests directory and the data
directory by their hash; a digest of all the data in the key.  This
should mean that verifying the key "out of band" (e.g., over email)
is easy.  But practically it should not matter much.  Signing the
keys would probably be a better idea.  In any case, this hash is
checked when the key is opened by the authorise or list commands.

## Limitations

In the dropbox scenario, non-password protected keys will afford
only limited protection.  This is because even though the keys and
data are stored separately on dropbox, they will be in the same
place on a local computer; if that computer is lost then the only
thing preventing an attacker recovering the data is security
through obscuritty (the data would appear to be random junk but
they will be able to run your analysis scripts as easily as you
can).  Password protected keys will improve this situation
considerably as without a password the data cannot be recovered.

The data and the key to encrypt it are not encrypted during a
running R session.  R allows arbitrary modification of code at
runtime so this package provides no security from the point where
the data can be decrypted.  If your computer was compromised then
stealing the data while you are running R should be assumed to be
straightforward.

``` {r echo=FALSE, results="hide"}
unlink(data_dir, recursive=TRUE)
```
