---
title: "Data Encryption"
author: "Rich FitzJohn"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Encryption}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**The scenario:**

A group of people are working on a sensitive data set that for
practical reasons needs to be stored in a place that we're not 100%
happy with the security (e.g., Dropbox), or we're concerned that
files stored in plain text on users computers (e.g. laptops) may
lead to the data being compromised.

If the data can be stored encrypted but everyone in the group can
still read and write the data then we've improved the situation
somewhat.  But organising for everyone to get a copy of the key to
decrypt the data files is non-trivial.  The workflow described here
aims to simplify this procedure using lower-level functions in the
`cyphr` package.

The general procedure is this:

1. A person will set up a set of personal keys and a key for the
data.  The data key will be encrypted with their personal key so
they have access to the data but nobody else does.  At this point
the data can be encrypted.

2. Additional users set up personal keys and request access to the
data.  Anyone with access to the data can grant access to anyone
else.

Before doing any of this, everyone needs to have ssh keys set up.
By default the package will use your ssh keys found at "~/.ssh";
see the main package vignette for how to use this.

For clarity here we will generate two sets of key pairs for two
actors Alice and Bob:
``` {r }
path_key_alice <- cyphr::ssh_keygen(password = FALSE)
path_key_bob <- cyphr::ssh_keygen(password = FALSE)
```

These would ordinarily be on different machines (nobody has access
to anyone else's private key) and they would be password protected.
In the function calls below, all the `path_user` arguments would be
omitted.

We'll store data in the directory `data`; at present there is
nothing there (this is in a temporary directory for compliance with
CRAN policies but would ordinarily be somewhere persistent and
under version control ideally).
``` {r }
data_dir <- file.path(tempdir(), "data")
dir.create(data_dir)
dir(data_dir)
```

**First**, create a personal set of keys.  These will be shared
across all projects and stored away from the data.  Ideally one
would do this with `ssh-keygen` at the command line, following one
of the many guides available.  A utility function `ssh_keygen`
(which simply calls `ssh-keygen` for you) is available in this
package though.  You will need to generate a key on each computer
you want access from.  Don't copy the key around.  If you lose your
user key you will lose access to the data!

**Second**, create a key for the data and encrypt that key with
your personal key.  Note that the data key is never stored directly
- it is always stored encrypted by a personal key.
``` {r }
cyphr::data_admin_init(data_dir, path_user = path_key_alice)
```

The data key is very important.  If it is deleted, then the data
cannot be decrypted.  So do not delete the directory
`data_dir/.cyphr`!  Ideally add it to your version control
system so that it cannot be lost.  Of course, if you're working in
a group, there are multiple copies of the data key (each encrypted
with a different person's personal key) which reduces the chance of
total loss.

This command can be run multiple times safely; if it detects it has
been rerun and the data key will not be regenerated.
``` {r }
cyphr::data_admin_init(data_dir, path_user = path_key_alice)
```

**Third**, you can add encrypted data to the directory (or to
anywhere really).  When run, `cyphr::config_data` will verify
that it can actually decrypt things.
``` {r }
key <- cyphr::data_key(data_dir, path_user = path_key_alice)
```

This object can be used with all the `cyphr` functions (see the
"cyphr" vignette; `vignette("cyphr")`)
``` {r }
filename <- file.path(data_dir, "iris.rds")
cyphr::encrypt(saveRDS(iris, filename), key)
dir(data_dir)
```

The file is encrypted and so cannot be read with `readRDS`:
``` {r error = TRUE}
readRDS(filename)
```

But we can decrypt and read it:
``` {r }
head(cyphr::decrypt(readRDS(filename), key))
```

**Fourth**, have someone else join in.  Recall that to simulate
another person here, I'm going to pass an argument `path_user =
path_key_bob` though to the functions.  This contains the path to
"Bob"'s ssh keypair.  If run on an actually different computer this
would not be needed; this is just to simulate two users in a single
session for this vignette (see minimal example below where this is
simulated).  Again, typically this user would also not use the
`cyphr::ssh_keygen` function but use the `ssh-keygen` command from
their shell.

We're going to assume that the user can read and write to the data.
This is the case for my use case where the data are stored on
dropbox and will be the case with GitHub based distribution, though
there would be a pull request step in here.

This user cannot read the data, though trying to will print a
message explaining how you might request access:
``` {r error = TRUE}
key_bob <- cyphr::data_key(data_dir, path_user = path_key_bob)
```

But `bob` is your collaborator and needs access!  What they need
to do is run:
``` {r }
cyphr::data_request_access(data_dir, path_user = path_key_bob)
```

(again, ordinarily you would not need the `bob` bit here)

The user should the send an email to someone with access and quote
the hash in the message above.

**Fifth**, back on the first computer we can authorise the second
user.  First, see who has requested access:
``` {r }
req <- cyphr::data_admin_list_requests(data_dir)
req
```

We can see the same hash here as above (``r names(req)[[1]]``)

...and then grant access to them with the
`cyphr::data_admin_authorise` function.
``` {r }
cyphr::data_admin_authorise(data_dir, yes = TRUE, path_user = path_key_alice)
```

If you do not specify `yes = TRUE` will prompt for confirmation at
each key added.

This has cleared the request queue:
``` {r }
cyphr::data_admin_list_requests(data_dir)
```

and added it to our set of keys:
``` {r }
cyphr::data_admin_list_keys(data_dir)
```

**Finally**, as soon as the authorisation has happened, the user
can encrypt and decrypt files:
``` {r }
key_bob <- cyphr::data_key(data_dir, path_user = path_key_bob)
head(cyphr::decrypt(readRDS(filename), key_bob))
```

## Minimal example

As above, but with less discussion:

``` {r echo = FALSE, results = "hide"}
unlink(data_dir, recursive = TRUE)
dir.create(data_dir)
```

Setup, on alice's computer computer:
``` {r }
cyphr::data_admin_init(data_dir, path_user = path_key_alice)
```

Get the data key key:
``` {r }
key <- cyphr::data_key(data_dir, path_user = path_key_alice)
```

Encrypt a file:
``` {r }
cyphr::encrypt(saveRDS(iris, filename), key)
```

Request access, on Bob's computer computer:
``` {r }
hash <- cyphr::data_request_access(data_dir, path_user = path_key_bob)
```

Alice authorises this request::
``` {r }
cyphr::data_admin_authorise(data_dir, yes = TRUE, path_user = path_key_alice)
```

Bob can get the data key:
``` {r }
key <- cyphr::data_key(data_dir, path_user = path_key_bob)
```

Bob can read the secret data:
``` {r }
head(cyphr::decrypt(readRDS(filename), key))
```

## Details & disclosure

Encryption does not work through security through obscurity; it
works because we can rely on the underlying maths enough to be open
about how things are stored and where.

Most encryption libraries require some degree of security in
the underlying software.  Because of the way R works this is very
difficult to guarantee; it is trivial to rewrite code in running
packages to skip past verification checks.  So this package is
_not_ designed to (or able to) avoid exploits in your running code;
an attacker could intercept your private keys, the private key to
the data, or skip the verification checks that are used to make
sure that the keys you load are what they say they are.  However,
the _data_ are safe; only people who have keys to the data will be
able to read it.

`cyphr` uses two different encryption algorithms; it uses RSA
encryption via the `openssl` package for user keys, because there
is a common file format for these keys so it makes user
configuration easier.  It uses the modern sodium package (and
through that the libsodium library) for data encryption because it
is very fast and simple to work with.  This does leave two possible
points of weakness as a vulnerability in either of these libraries
could lead to an exploit that could allow decryption of your data.

Each user has a public/private key pair.  Typically this is in
`~/.ssh/id_rsa.pub` and `~/.ssh/id_rsa`, and if found these will be
used.  Alternatively the location of the keypair can be stored
elsewhere and pointed at with the `USER_KEY` or `USER_PUBKEY`
environment variables.  The key may be password protected (and this
is recommended!) and the password will be requested without ever
echoing it to the terminal.

The data directory has a hidden directory `.cyphr` in it.
``` {r }
dir(data_dir, all.files = TRUE, no.. = TRUE)
```

This does not actually need to be stored with the data but it
makes sense to (there are workflows where data is stored remotely
where storing this directory might make sense).  This directory
contains a number of files; one for each person who has access to
the data.
``` {r }
dir(file.path(data_dir, ".cyphr"))
names(cyphr::data_admin_list_keys(data_dir))
```

(the file `test` is a small file encrypted with the data key used
to verify everything is working OK).

Each file is stored in RDS format and is a list with elements:

* user: the reported user name of the person who created request for data
* host: the reported computer name
* date: the time the request was generated
* pub: the RSA public key of the user
* signature: the signature of the contents of "user", "host",
  "date", "pub".  This ensures that the data have not been changed
  since they were created.
* key: the data key, encrypted with the user key.  Without the
  private key, this cannot be used.  With the user's private key
  this can be used to generate the symmetric key to the data.

``` {r }
h <- names(cyphr::data_admin_list_keys(data_dir))[[1]]
readRDS(file.path(data_dir, ".cyphr", h))
```

You can see that the hash of the public key is the same as name of
the stored file here (which is used to prevent collisions when
multiple people request access at the same time).
``` {r }
h
```

When a request is posted it is an RDS file with all of the above
except for the `key` element, which is added during authorisation.

(Note that the verification relies on the package code not being
attacked, and given R's highly dynamic nature an attacker could
easily swap out the definition for the verification function with
something that always returns `TRUE`.)

When an authorised user creates the `data_key` object (which
allows decryption of the data) `secret` will:

* read their private user key (probably from `~/.ssh/id_rsa`)
* read the encrypted data key from the data directory (the `$key`
  element from the list above).
* decrypt this data key using their user key to yield the the data
  symmetric key.

## Limitations

In the Dropbox scenario, non-password protected keys will afford
only limited protection.  This is because even though the keys and
data are stored separately on Dropbox, they will be in the same
place on a local computer; if that computer is lost then the only
thing preventing an attacker recovering the data is security
through obscurity (the data would appear to be random junk but
they will be able to run your analysis scripts as easily as you
can).  Password protected keys will improve this situation
considerably as without a password the data cannot be recovered.

The data is not encrypted during a running R session.  R allows
arbitrary modification of code at runtime so this package provides
no security from the point where the data can be decrypted.  If
your computer was compromised then stealing the data while you are
running R should be assumed to be straightforward.

``` {r echo = FALSE, results = "hide"}
unlink(c(data_dir, path_key_alice, path_key_bob), recursive = TRUE)
```
